Finite-state morphological transducers for three Kypchak languages

### 2014-03-18 kipchak-lrec-2013/kipchak-paper.tex
better map, 2014 stylesheet, introduction
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@50925 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{lrec2012}

@@

\author{Jonathan North Washington \\
Departments of Linguistics and Central Eurasian Studies\\
Indiana University\\
Bloomington, IN 47405 (USA)\\
\texttt{jonwashi@indiana.edu} \and
Ilnar Salimzyanov  \\
Kazan Federal University \\
Kazan, Republic of Tatarstan\\
(Russian Federation) \\
\texttt{ilnar.salimzyan@gmail.com} \and 
Francis M. Tyers\\
HSL-fakultehta, \\
UiT Norgga árktalaš universitehta, \\
9019 Romsa (Norway)\\
\texttt{francis.tyers@uit.no} 
}

@@

This paper describes the development of morphological transducers for three closly related languages: Kazakh, Tatar, and Kumyk.

@@

The transducers for these languages

@@

A finite-state transducer which models the morphology of a language has a wide variety of usages. For example,
for creating proofing tools such as spellcheckers and grammarcheckers, and for machine translation.

\section{Languages}

The three languages for which transducers were developed belong to the Northwestern branch of the Turkic family, which is often referred to as the Kypchak branch.  This branch can be divided into three subbranches.  Kumyk is a member of the Western Kypchak group, Tatar is a member of the Northern Kypchak group, and Kazakh is a member of the Southern Kypchak group \citep[82-83]{histofturkic}.  As such, each of these three languages represents one of the three branches of Kypchak.  The geographic distribution of the languages is shown in map \ref{map1}.

@@

\section{Background}
\subsection{Morphological transducers, previous work}
%FIXME: need a source, maybe something by Tommi?
The objective of a morphological transducer is twofold: firstly to take surface forms (e.g., алдым) and generate all possible lexical forms, and secondly to take lexical forms (e.g.,  ал{\tt {\small <v><tv><ifi><p1><sg>}}, алд{\tt {\small <n><px1sg><nom>}}, etc.) and generate one or more surface forms.  As they are implemented as finite-state transducers, they are reversible by default. For more information on usin finite-state transducers for morphological analysis and generation, the reader is referred to \cite{beesley2003}.

@@

\section{Methodology}
% cite Hengeveld (1992) ?

@@

\section{Evaluation}

@@

\section{Conclusions}

>>>>

\usepackage{lrec2014}

@@

%\author{Jonathan North Washington \\
%Departments of Linguistics and Central Eurasian Studies\\
%Indiana University\\
%Bloomington, IN 47405 (USA)\\
%\texttt{jonwashi@indiana.edu} \and
%Ilnar Salimzyanov  \\
%Kazan Federal University \\
%Kazan, Republic of Tatarstan\\
%(Russian Federation) \\
%\texttt{ilnar.salimzyan@gmail.com} \and 
%Francis M. Tyers\\
%5HSL-fakultehta, \\
%UiT Norgga árktalaš universitehta, \\
%9019 Romsa (Norway)\\
%\texttt{francis.tyers@uit.no} 
%}

@@

This paper describes the development of free/open-source morphological transducers for three closly related languages: Kazakh, Tatar, and Kumyk.  Morphological transducers are computational models of the languages' morphology, and are used to output morphological analyses from word forms and vice-versa.

These transducers were all developed as part of the Apertium project, which is aimed at creating rule-based MT systems for lesser resourced languages.  As such, the transducers were developed with the intent that they be used as morphological analysers and generators in RBMT systems.  However, this sort of finite-state transducer has a wide variety of usages, including for creating proofing tools such as spellcheckers and grammarcheckers.

%the potential uses of morphological transducers is much broader; for example, we have found that they are easily converted to spell checkers.

All three transducers were designed as ``ports'' of Apertium's Kyrgyz morphological transducer \citep{washington2012}.  The Tatar and Kazakh transducers are used in a release-quality Kazakh-Tatar MT system \citep{salimzyanov2013}.

@@

%The transducers for these languages

@@

This paper overviews the languages (§\ref{sec:lgs}), provides background on these morphological transducers and related previous work (§\ref{sec:bg}), describes the development effort and contents of the transducers (§\ref{sec:methodology}), evaluates the effectiveness of these transducers (§\ref{sec:evaluation}), and summarises the findings (§\ref{sec:conclusions}).

\section{Languages}\label{sec:lgs}

The three languages for which transducers were developed belong to the Northwestern branch of the Turkic family, which is often referred to as the Kypchak branch.  This branch can be divided into three subbranches.\footnote{It is the belief of the authors of this paper that Kyrgyz constitutes a fourth branch.}  Kumyk is a member of the Western Kypchak group, Tatar is a member of the Northern Kypchak group, and Kazakh is a member of the Southern Kypchak group \citep[82-83]{histofturkic}.  As such, each of these three languages represents one of the three branches of Kypchak.  The geographic distribution of the languages is shown in map \ref{map1}.

@@

\section{Background}\label{sec:bg}
\subsection{Morphological transducers, previous work}
%FIXME: need a source, maybe something by Tommi?
The objective of a morphological transducer is twofold: firstly to take surface forms (e.g., алдым) and generate all possible lexical forms, and secondly to take lexical forms (e.g.,  ал{\tt {\small <v><tv><ifi><p1><sg>}}, алд{\tt {\small <n><px1sg><nom>}}, etc.) and generate one or more surface forms.  As they are implemented as finite-state transducers, they are reversible by default. For more information on usin finite-state transducers for morphological analysis and generation, the reader is referred to \cite{beesley2003}.

@@

\section{Methodology}\label{sec:methodology}
% cite Hengeveld (1992) ?

@@

\section{Evaluation}\label{sec:evaluation}

@@

\section{Conclusions}\label{sec:conclusions}
